---
title: "Seminar Presentation"
author: "Julie Wilson"
format: html
editor: visual
---

## Load Packages

```{r load_packages}

library("pacman")
library("tidyverse")
library("maps")
library(dplyr)
library(did)
library(bacondecomp)
library(fixest)
library(modelsummary)
library(infer)
library(ggplot2)
library(car)
library(plm)
library(readr)
library(broom)
library(kableExtra)

library(corrplot)

library(stargazer)

knitr::opts_knit$set(root.dir = "C:/Users/julie/Downloads/Seminar research/even more r/Final_seminar")
```

## Import Data

```{r data_import}

df_all <- read.csv("C:/Users/julie/Downloads/Seminar research/even more r/Final_seminar/all_data.csv")
names(df_all)
nrow(df_all)
```

### Checking for missing data

```{r missing_data_check}

# Check for missing data in df_all
sum(is.na(df_all))
colSums(is.na(df_all))
```

### creating policy groups

```{r creating_policy_groups}
state_min <- tapply(df_all$gestational_limit_weeks, df_all$State, min)
df_all$policy_group <- ifelse(state_min[df_all$State] == 0, "Total Ban",
                       ifelse(state_min[df_all$State] <= 6, "Severely Restrictive",
                       ifelse(state_min[df_all$State] <= 12, "Moderately Restrictive",
                       ifelse(state_min[df_all$State] <= 18, "Moderately Permissive",
                       "Permissive"))))

# Check
table(df_all$policy_group)
state_policy_map <- unique(df_all[, c("State", "policy_group")])
split(state_policy_map$State, state_policy_map$policy_group)


```

```{r mod perm}

library(dplyr)
library(ggplot2)
library(maps)

class(df_all)

df_all$policy_group <- ifelse(state_min[df_all$State] == 0, "Total Ban",
                       ifelse(state_min[df_all$State] <= 6, "Severely Restrictive",
                       ifelse(state_min[df_all$State] <= 12, "Moderately Restrictive",
                       ifelse(state_min[df_all$State] <= 18, "Moderately Permissive",
                       "Permissive"))))
class(df_all)
str(df_all)

df_all <- df_all %>%
  mutate(policy_group = case_when(
    state_min[State] == 0 ~ "Total Ban",
    state_min[State] <= 6 ~ "Severely Restrictive",
    state_min[State] <= 12 ~ "Moderately Restrictive",
    state_min[State] <= 18 ~ "Moderately Permissive",
    TRUE ~ "Permissive"
  ))

df_all <- df_all%>%
  mutate(
    State = trimws(toupper(State)), 
    region = tolower(state.name[match(State, state.abb)])
  )

df_all %>% filter(is.na(region)) %>% select(State)

library(ggplot2)
library(maps)


ggplot() +

  # base US map
  geom_map(data = us_map, map = us_map,
           aes(map_id = region),
           fill = "gray90", color = "black") +

  # your highlighted group (example)
  geom_map(data = df_all %>% filter(policy_group == "Moderately Permissive"),
           map = us_map,
           aes(map_id = region),
           fill = "#91bfdb", color = "black") +
  theme_void()+

  # ALL state labels
  geom_text(
    data = state_centers,
    aes(x = lon, y = lat, label = abb),
    size = 2
  ) +
  

  expand_limits(x = us_map$long, y = us_map$lat) +
  labs(title = "Moderately Permissive States")





```

```{r total ban map}

ggplot() +
  
  # base US map
  geom_map(data = us_map, map = us_map,
           aes(map_id = region),
           fill = "gray90", color = "black") +
  
  # highlight Total Ban states
  geom_map(
    data = df_all %>% filter(policy_group == "Total Ban"),
    map = us_map,
    aes(map_id = region),
    fill = "#d73027", color = "black"
  ) +
  
  # state abbreviations
  geom_text(
    data = state_centers,
    aes(x = lon, y = lat, label = abb),
    size = 2
  ) +
  
  expand_limits(x = us_map$long, y = us_map$lat) +
  theme_void() +
  labs(title = "Total Ban States")
```

```{r sev restricted map}

ggplot() +
  
  # base US map
  geom_map(data = us_map, map = us_map,
           aes(map_id = region),
           fill = "gray90", color = "black") +
  
  # highlight Severely Restrictive states
  geom_map(
    data = df_all %>% filter(policy_group == "Severely Restrictive"),
    map = us_map,
    aes(map_id = region),
    fill = "#fc8d59", color = "black"
  ) +
  
  # state abbreviations
  geom_text(
    data = state_centers,
    aes(x = lon, y = lat, label = abb),
    size = 2
  ) +
  
  expand_limits(x = us_map$long, y = us_map$lat) +
  theme_void() +
  labs(title = "Severely Restrictive States")
```

```{r mod restrict map}

ggplot() +
  
  # base US map
  geom_map(data = us_map, map = us_map,
           aes(map_id = region),
           fill = "gray90", color = "black") +
  
  # highlight Moderately Restrictive states
  geom_map(
    data = df_all %>% filter(policy_group == "Moderately Restrictive"),
    map = us_map,
    aes(map_id = region),
    fill = "#fee090", color = "black"
  ) +
  
  # state abbreviations
  geom_text(
    data = state_centers,
    aes(x = lon, y = lat, label = abb),
    size = 2
  ) +
  
  expand_limits(x = us_map$long, y = us_map$lat) +
  theme_void() +
  labs(title = "Moderately Restrictive States")
```

```{r perm map}

ggplot() +
  
  # base US map
  geom_map(data = us_map, map = us_map,
           aes(map_id = region),
           fill = "gray90", color = "black") +
  
  # highlight Permissive states
  geom_map(
    data = df_all %>% filter(policy_group == "Permissive"),
    map = us_map,
    aes(map_id = region),
    fill = "#4575b4", color = "black"
  ) +
  
  # state abbreviations
  geom_text(
    data = state_centers,
    aes(x = lon, y = lat, label = abb),
    size = 2
  ) +
  
  expand_limits(x = us_map$long, y = us_map$lat) +
  theme_void() +
  labs(title = "Permissive States")
```

## Descriptive stats

-    Abortion Rate per 1,000 women 15-44

<!-- -->

-   TANF Applications per 1,000 women 15-44

<!-- -->

-   TANF Caseloads per 1,000 women 15-44

<!-- -->

-   SNAP Participation per 1,000 women 15-44

<!-- -->

-   Single Mother Median Monthly Income

<!-- -->

-   Unemployment Rate

<!-- -->

-   Minimum Wage

<!-- -->

-   Childcare Cost Burden

<!-- -->

-   Medicaid Abortion Funding Restrictions

```{r descriptive_stats}

names(df_all)[grepl("tanf|snap|abort|income|unemploy|wage|child", 
                     names(df_all), ignore.case = TRUE)]

desc_stats <- df_all %>%
  filter(!is.na(policy_group)) %>%
  mutate(period = if_else(months_since_dobbs < 0, "Pre-Dobbs", "Post-Dobbs")) %>%
  group_by(policy_group, period) %>%
  summarize(
    mean_abortion_rate     = mean(abortions_per_1000_women_ages_1544,          na.rm = TRUE),
    mean_tanf_apps         = mean(tanf._apps_per_1000,                         na.rm = TRUE),
    mean_tanf_caseloads    = mean(tanf_one_parent_per1000_women_1544,          na.rm = TRUE),
    mean_snap              = mean(snap_participation_rate,                      na.rm = TRUE),
    mean_single_mom_income = mean(single_mothers_median_monthly_income,        na.rm = TRUE),
    mean_unemployment      = mean(state_unemployment_rate,                     na.rm = TRUE),
    mean_min_wage          = mean(minimum_wage,                                na.rm = TRUE),
    mean_childcare         = mean(infant_child_care_cost_burden,               na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  arrange(policy_group, period)

desc_stats$policy_group <- factor(desc_stats$policy_group, 
                                   levels = c("Total Ban", 
                                              "Severely Restrictive", 
                                              "Moderately Restrictive", 
                                              "Moderately Permissive", 
                                              "Permissive"))

desc_stats <- desc_stats %>% arrange(policy_group, period)

# Then re-split
pre_dobbs <- desc_stats %>% filter(period == "Pre-Dobbs") %>% select(-period)
post_dobbs <- desc_stats %>% filter(period == "Post-Dobbs") %>% select(-period)
knitr::kable(desc_stats,
  col.names = c("Policy Group", "Period",
                "Abortion Rate", "TANF Applications",
                "TANF Caseloads", "SNAP Participation",
                "Single Mom Monthly Income", "Unemployment Rate",
                "Minimum Wage", "Childcare Cost Burden"),
  caption = "Descriptive Statistics by Policy Group Pre and Post Dobbs") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = TRUE) %>%
  kableExtra::row_spec(which(desc_stats$period == "Pre-Dobbs"),
                       background = "#fff3cd") %>%  # yellow = Pre-Dobbs
  kableExtra::row_spec(which(desc_stats$period == "Post-Dobbs"),
                       background = "#cce5ff")       # blue = Post-Dobbs

# Split into two tables
pre_dobbs <- desc_stats %>% filter(period == "Pre-Dobbs") %>% select(-period)
post_dobbs <- desc_stats %>% filter(period == "Post-Dobbs") %>% select(-period)

# Pre-Dobbs table
knitr::kable(pre_dobbs,
  col.names = c("Policy Group",
                "Abortion Rate", "TANF Apps",
                "TANF Caseloads", "SNAP Participation",
                "Single Mom Income", "Unemployment Rate",
                "Minimum Wage", "Childcare Burden"),
  caption = "Pre-Dobbs Descriptive Statistics by Policy Group") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = TRUE) %>%
  kableExtra::row_spec(0, background = "#fff3cd", bold = TRUE)  # yellow header

# Post-Dobbs table
knitr::kable(post_dobbs,
  col.names = c("Policy Group",
                "Abortion Rate", "TANF Apps",
                "TANF Caseloads", "SNAP Participation",
                "Single Mom Income", "Unemployment Rate",
                "Minimum Wage", "Childcare Burden"),
  caption = "Post-Dobbs Descriptive Statistics by Policy Group") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = TRUE) %>%
  kableExtra::row_spec(0, background = "#cce5ff", bold = TRUE)  # blue header


```

Since adjusting policy groups, medicaid abortion funding restrictions and medicaid expansion was dropped because it showed no change pre and post dobbs.

### Testing for multicollinearity

```{r multicollinearity_check}
# Select the variables you want to test
corr_vars <- df_all %>%
  select(abortions_per_1000_women_ages_1544,
         tanf._apps_per_1000,
         tanf_one_parent_per1000_women_1544,
         snap_participation_rate,
         single_mothers_median_monthly_income,
         state_unemployment_rate,
         minimum_wage,
         infant_child_care_cost_burden,
         medicaid_abortion_funding_restrictions,
         medicaid_expansion_status) %>%
  na.omit()

# Correlation matrix
cor_matrix <- cor(corr_vars)
round(cor_matrix, 2)

# Visual correlation matrix (optional but easier to read)
library(corrplot)
corrplot(cor_matrix, method = "color", type = "upper",
         tl.cex = 0.7, tl.col = "black",
         addCoef.col = "black", number.cex = 0.6)
```

## Parallel Trends Credibility Check

-   Fundamental identification assumption for Difference-in-Differences (DiD) model, which posits that in the absence of the *Dobbs* decision, treatment and control states would have followed similar economic trajectories.

##### TANF Caseloads

```{r parallel_trends_check}
df_parallel <- df_all %>%
  filter(!is.na(policy_group),
         months_since_dobbs >= -3,
         months_since_dobbs <= 0) %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(
    mean_tanf = mean(tanf_one_parent_per1000_women_1544, na.rm = TRUE),
    se_tanf   = sd(tanf_one_parent_per1000_women_1544, na.rm = TRUE) /
                sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(
    ci_low  = mean_tanf - 1.96 * se_tanf,
    ci_high = mean_tanf + 1.96 * se_tanf,
    policy_group = factor(policy_group,
                          levels = c("Total Ban",
                                     "Severely Restrictive",
                                     "Moderately Restrictive",
                                     "Moderately Permissive",
                                     "Permissive"))
  )

ggplot(df_parallel,
       aes(x = months_since_dobbs, y = mean_tanf,
           color = policy_group, fill = policy_group)) +
  geom_ribbon(aes(ymin = ci_low, ymax = ci_high),
              alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0,
             linetype = "dashed",
             color = "red",
             linewidth = 0.8) +
  annotate("text",
           x = -0.1, y = Inf,
           label = "Dobbs Decision",
           hjust = 1, vjust = 2,
           color = "red",
           size = 3.5) +
  scale_color_manual(values = c(
    "Total Ban"              = "#d73027",   # red
    "Severely Restrictive"   = "#fc8d59",   # orange
    "Moderately Restrictive" = "#fee090",   # yellow
    "Moderately Permissive"  = "#91bfdb",   # light blue
    "Permissive"             = "#4575b4"    # dark blue
  )) +
  scale_fill_manual(values = c(
    "Total Ban"              = "#d73027",
    "Severely Restrictive"   = "#fc8d59",
    "Moderately Restrictive" = "#fee090",
    "Moderately Permissive"  = "#91bfdb",
    "Permissive"             = "#4575b4"
  )) +
  scale_x_continuous(breaks = c(-3, -2, -1, 0)) +
  labs(
    title    = "Parallel Trends Check - Pre-Dobbs TANF Caseloads",
    x        = "Months Since Dobbs (0 = June 2022)",
    y        = "TANF Cases per 1,000 Women (15-44)",
    color    = "Policy Group",
    fill     = "Policy Group"
  ) +
  theme_minimal() +
  theme(plot.margin = margin(10, 10, 10, 40))
```

##### Abortion Rate and Snap Participation

```{r parallel_trends_abortion_rate}
# Parallel Trends Check - Abortion Rate
df_parallel_abort <- df_all %>%
  filter(!is.na(policy_group),
         months_since_dobbs >= -3,
         months_since_dobbs <= 0) %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(
    mean_outcome = mean(abortions_per_1000_women_ages_1544, na.rm = TRUE),
    se_outcome   = sd(abortions_per_1000_women_ages_1544, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(
    ci_low  = mean_outcome - 1.96 * se_outcome,
    ci_high = mean_outcome + 1.96 * se_outcome,
    policy_group = factor(policy_group,
                          levels = c("Total Ban", "Severely Restrictive",
                                     "Moderately Restrictive", "Moderately Permissive",
                                     "Permissive"))
  )

# Parallel Trends Check - SNAP
df_parallel_snap <- df_all %>%
  filter(!is.na(policy_group),
         months_since_dobbs >= -3,
         months_since_dobbs <= 0) %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(
    mean_outcome = mean(snap_participation_rate, na.rm = TRUE),
    se_outcome   = sd(snap_participation_rate, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(
    ci_low  = mean_outcome - 1.96 * se_outcome,
    ci_high = mean_outcome + 1.96 * se_outcome,
    policy_group = factor(policy_group,
                          levels = c("Total Ban", "Severely Restrictive",
                                     "Moderately Restrictive", "Moderately Permissive",
                                     "Permissive"))
  )

# Plot function to avoid repeating code
plot_parallel <- function(data, title, ylab) {
  ggplot(data, aes(x = months_since_dobbs, y = mean_outcome,
                   color = policy_group, fill = policy_group)) +
    geom_ribbon(aes(ymin = ci_low, ymax = ci_high), alpha = 0.2, color = NA) +
    geom_line(linewidth = 1) +
    geom_point(size = 3) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) +
    annotate("text", x = -0.1, y = Inf, label = "Dobbs Decision",
             hjust = 1, vjust = 2, color = "red", size = 3.5) +
    scale_color_manual(values = c(
      "Total Ban"              = "#d73027",
      "Severely Restrictive"   = "#fc8d59",
      "Moderately Restrictive" = "#fee090",
      "Moderately Permissive"  = "#91bfdb",
      "Permissive"             = "#4575b4"
    )) +
    scale_fill_manual(values = c(
      "Total Ban"              = "#d73027",
      "Severely Restrictive"   = "#fc8d59",
      "Moderately Restrictive" = "#fee090",
      "Moderately Permissive"  = "#91bfdb",
      "Permissive"             = "#4575b4"
    )) +
    scale_x_continuous(breaks = c(-3, -2, -1, 0)) +
    labs(title = title, x = "Months Since Dobbs (0 = June 2022)",
         y = ylab, color = "Policy Group", fill = "Policy Group") +
    theme_minimal() +
    theme(plot.margin = margin(10, 10, 10, 40))
}

# Generate plots
plot_parallel(df_parallel_abort,
              "Parallel Trends Check - Pre-Dobbs Abortion Rate",
              "Abortions per 1,000 Women (15-44)")

plot_parallel(df_parallel_snap,
              "Parallel Trends Check - Pre-Dobbs SNAP Participation",
              "SNAP Participation Rate")
```

##### TANF Apps

```{r tanf_apps_prallel}
df_parallel_tanf_apps <- df_all %>%
  filter(!is.na(policy_group),
         months_since_dobbs >= -3,
         months_since_dobbs <= 0) %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(
    mean_outcome = mean(tanf._apps_per_1000, na.rm = TRUE),
    se_outcome   = sd(tanf._apps_per_1000, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(
    ci_low  = mean_outcome - 1.96 * se_outcome,
    ci_high = mean_outcome + 1.96 * se_outcome,
    policy_group = factor(policy_group,
                          levels = c("Total Ban", "Severely Restrictive",
                                     "Moderately Restrictive", "Moderately Permissive",
                                     "Permissive"))
  )

plot_parallel(df_parallel_tanf_apps,
              "Parallel Trends Check - Pre-Dobbs TANF Applications",
              "TANF Applications per 1,000 Women (15-44)")
```

### Variables by State Policy Group

```{r by_state_policy_group}
# Abortion Rate - Full Timeline
df_all %>%
  filter(!is.na(policy_group)) %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(mean_outcome = mean(abortions_per_1000_women_ages_1544, na.rm = TRUE), 
            .groups = "drop") %>%
  mutate(policy_group = factor(policy_group,
                               levels = c("Total Ban", "Severely Restrictive",
                                          "Moderately Restrictive", "Moderately Permissive",
                                          "Permissive"))) %>%
  ggplot(aes(x = months_since_dobbs, y = mean_outcome, color = policy_group)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = 0.5, y = Inf, label = "Dobbs Decision",
           hjust = 0, vjust = 2, color = "red", size = 3.5) +
  scale_color_manual(values = c(
    "Total Ban"              = "#d73027",
    "Severely Restrictive"   = "#fc8d59",
    "Moderately Restrictive" = "#fee090",
    "Moderately Permissive"  = "#91bfdb",
    "Permissive"             = "#4575b4"
  )) +
  labs(title = "Abortion Rate by Policy Group Over Time",
       x = "Months Since Dobbs (0 = June 2022)",
       y = "Abortions per 1,000 Women (15-44)",
       color = "Policy Group") +
  theme_minimal() +
  theme(plot.margin = margin(10, 10, 10, 40))

 # SNAP Participation - Full Timeline
df_all %>%
  filter(!is.na(policy_group)) %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(mean_outcome = mean(snap_participation_rate, na.rm = TRUE), 
            .groups = "drop") %>%
  mutate(policy_group = factor(policy_group,
                               levels = c("Total Ban", "Severely Restrictive",
                                          "Moderately Restrictive", "Moderately Permissive",
                                          "Permissive"))) %>%
  ggplot(aes(x = months_since_dobbs, y = mean_outcome, color = policy_group)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = 0.5, y = Inf, label = "Dobbs Decision",
           hjust = 0, vjust = 2, color = "red", size = 3.5) +
  scale_color_manual(values = c(
    "Total Ban"              = "#d73027",
    "Severely Restrictive"   = "#fc8d59",
    "Moderately Restrictive" = "#fee090",
    "Moderately Permissive"  = "#91bfdb",
    "Permissive"             = "#4575b4"
  )) +
  labs(title = "SNAP Participation by Policy Group Over Time",
       x = "Months Since Dobbs (0 = June 2022)",
       y = "SNAP Participation Rate",
       color = "Policy Group") +
  theme_minimal() +
  theme(plot.margin = margin(10, 10, 10, 40))

# TANF Caseloads - Full Timeline
df_all %>%
  filter(!is.na(policy_group)) %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(mean_outcome = mean(tanf_one_parent_per1000_women_1544, na.rm = TRUE), 
            .groups = "drop") %>%
  mutate(policy_group = factor(policy_group,
                               levels = c("Total Ban", "Severely Restrictive",
                                          "Moderately Restrictive", "Moderately Permissive",
                                          "Permissive"))) %>%
  ggplot(aes(x = months_since_dobbs, y = mean_outcome, color = policy_group)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = 0.5, y = Inf, label = "Dobbs Decision",
           hjust = 0, vjust = 2, color = "red", size = 3.5) +
  scale_color_manual(values = c(
    "Total Ban"              = "#d73027",
    "Severely Restrictive"   = "#fc8d59",
    "Moderately Restrictive" = "#fee090",
    "Moderately Permissive"  = "#91bfdb",
    "Permissive"             = "#4575b4"
  )) +
  labs(title = "TANF Caseloads by Policy Group Over Time",
       x = "Months Since Dobbs (0 = June 2022)",
       y = "TANF Cases per 1,000 Women (15-44)",
       color = "Policy Group") +
  theme_minimal() +
  theme(plot.margin = margin(10, 10, 10, 40))
```

```{r tanf app by group}
# TANF Applications - CPI Adjusted, Indexed to Pre-Dobbs Baseline
df_all %>%
  filter(!is.na(policy_group)) %>%
  mutate(outcome_adjusted = tanf._apps_per_1000 / CPI * 100) %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(mean_outcome = mean(outcome_adjusted, na.rm = TRUE), .groups = "drop") %>%
  group_by(policy_group) %>%
  mutate(baseline = mean(mean_outcome[months_since_dobbs < 0], na.rm = TRUE),
         mean_indexed = mean_outcome - baseline,
         policy_group = factor(policy_group,
                               levels = c("Total Ban", "Severely Restrictive",
                                          "Moderately Restrictive", "Moderately Permissive",
                                          "Permissive"))) %>%
  ggplot(aes(x = months_since_dobbs, y = mean_indexed, color = policy_group)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = 0.5, y = Inf, label = "Dobbs Decision",
           hjust = 0, vjust = 2, color = "red", size = 3.5) +
  scale_color_manual(values = c(
    "Total Ban"              = "#d73027",
    "Severely Restrictive"   = "#fc8d59",
    "Moderately Restrictive" = "#fee090",
    "Moderately Permissive"  = "#91bfdb",
    "Permissive"             = "#4575b4"
  )) +
  labs(title = "TANF Applications by State Policy Group - Pre and Post Dobbs",
       x = "Months Since Dobbs (0 = June 2022)",
       y = "Change in TANF Applications from Pre-Dobbs Baseline (CPI Adjusted)",
       color = "Policy Group") +
  theme_minimal() +
  theme(plot.margin = margin(10, 10, 10, 40))
```

```{r cpi_adjusted}
# Abortion Rate - CPI Adjusted, Indexed to Pre-Dobbs Baseline
df_all %>%
  filter(!is.na(policy_group)) %>%
  mutate(outcome_adjusted = abortions_per_1000_women_ages_1544 / CPI * 100) %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(mean_outcome = mean(outcome_adjusted, na.rm = TRUE), .groups = "drop") %>%
  group_by(policy_group) %>%
  mutate(baseline = mean(mean_outcome[months_since_dobbs < 0], na.rm = TRUE),
         mean_indexed = mean_outcome - baseline,
         policy_group = factor(policy_group,
                               levels = c("Total Ban", "Severely Restrictive",
                                          "Moderately Restrictive", "Moderately Permissive",
                                          "Permissive"))) %>%
  ggplot(aes(x = months_since_dobbs, y = mean_indexed, color = policy_group)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = 0.5, y = Inf, label = "Dobbs Decision",
           hjust = 0, vjust = 2, color = "red", size = 3.5) +
  scale_color_manual(values = c(
    "Total Ban"              = "#d73027",
    "Severely Restrictive"   = "#fc8d59",
    "Moderately Restrictive" = "#fee090",
    "Moderately Permissive"  = "#91bfdb",
    "Permissive"             = "#4575b4"
  )) +
  labs(title = "Abortion Rate by State Policy Group - Pre and Post Dobbs",
       x = "Months Since Dobbs (0 = June 2022)",
       y = "Change in Abortion Rate from Pre-Dobbs Baseline (CPI Adjusted)",
       color = "Policy Group") +
  theme_minimal() +
  theme(plot.margin = margin(10, 10, 10, 40))

# SNAP Participation - CPI Adjusted, Indexed to Pre-Dobbs Baseline
df_all %>%
  filter(!is.na(policy_group)) %>%
  mutate(outcome_adjusted = snap_participation_rate / CPI * 100) %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(mean_outcome = mean(outcome_adjusted, na.rm = TRUE), .groups = "drop") %>%
  group_by(policy_group) %>%
  mutate(baseline = mean(mean_outcome[months_since_dobbs < 0], na.rm = TRUE),
         mean_indexed = mean_outcome - baseline,
         policy_group = factor(policy_group,
                               levels = c("Total Ban", "Severely Restrictive",
                                          "Moderately Restrictive", "Moderately Permissive",
                                          "Permissive"))) %>%
  ggplot(aes(x = months_since_dobbs, y = mean_indexed, color = policy_group)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = 0.5, y = Inf, label = "Dobbs Decision",
           hjust = 0, vjust = 2, color = "red", size = 3.5) +
  scale_color_manual(values = c(
    "Total Ban"              = "#d73027",
    "Severely Restrictive"   = "#fc8d59",
    "Moderately Restrictive" = "#fee090",
    "Moderately Permissive"  = "#91bfdb",
    "Permissive"             = "#4575b4"
  )) +
  labs(title = "SNAP Participation by State Policy Group - Pre and Post Dobbs",
       x = "Months Since Dobbs (0 = June 2022)",
       y = "Change in SNAP Participation from Pre-Dobbs Baseline (CPI Adjusted)",
       color = "Policy Group") +
  theme_minimal() +
  theme(plot.margin = margin(10, 10, 10, 40))

# TANF Caseloads - CPI Adjusted, Indexed to Pre-Dobbs Baseline
df_all %>%
  filter(!is.na(policy_group)) %>%
  mutate(outcome_adjusted = tanf_one_parent_per1000_women_1544 / CPI * 100) %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(mean_outcome = mean(outcome_adjusted, na.rm = TRUE), .groups = "drop") %>%
  group_by(policy_group) %>%
  mutate(baseline = mean(mean_outcome[months_since_dobbs < 0], na.rm = TRUE),
         mean_indexed = mean_outcome - baseline,
         policy_group = factor(policy_group,
                               levels = c("Total Ban", "Severely Restrictive",
                                          "Moderately Restrictive", "Moderately Permissive",
                                          "Permissive"))) %>%
  ggplot(aes(x = months_since_dobbs, y = mean_indexed, color = policy_group)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = 0.5, y = Inf, label = "Dobbs Decision",
           hjust = 0, vjust = 2, color = "red", size = 3.5) +
  scale_color_manual(values = c(
    "Total Ban"              = "#d73027",
    "Severely Restrictive"   = "#fc8d59",
    "Moderately Restrictive" = "#fee090",
    "Moderately Permissive"  = "#91bfdb",
    "Permissive"             = "#4575b4"
  )) +
  labs(title = "TANF Caseloads by State Policy Group - Pre and Post Dobbs",
       x = "Months Since Dobbs (0 = June 2022)",
       y = "Change in TANF Caseloads from Pre-Dobbs Baseline (CPI Adjusted)",
       color = "Policy Group") +
  theme_minimal() +
  theme(plot.margin = margin(10, 10, 10, 40))
```

```{r panel_df_creation}

# Create panel data frame
df_panel <- pdata.frame(df_all, index = c("State", "year_month"))

# Verify it worked
exists("df_panel")
pdim(df_panel)

```

```{r model_existence}

df_panel$policy_group <- factor(df_panel$policy_group,
                                levels = c("Total Ban",
                                           "Severely Restrictive",
                                           "Moderately Restrictive",
                                           "Moderately Permissive",
                                           "Permissive"))

# TANF Applications
tanf_apps_ols      <- plm(tanf._apps_per_1000 ~ policy_group,
                          data = df_panel, model = "pooling")
tanf_apps_controls <- plm(tanf._apps_per_1000 ~ policy_group +
                            state_unemployment_rate + minimum_wage + CPI +
                            medicaid_abortion_funding_restrictions +
                            infant_child_care_cost_burden + RPP_all_items,
                          data = df_panel, model = "pooling")
tanf_apps_twfe     <- plm(tanf._apps_per_1000 ~ policy_group +
                            state_unemployment_rate + minimum_wage + CPI +
                            medicaid_abortion_funding_restrictions +
                            infant_child_care_cost_burden + RPP_all_items,
                          data = df_panel, model = "within", effect = "twoways")

# TANF Caseloads
tanf_cases_ols      <- plm(tanf_one_parent_per1000_women_1544 ~ policy_group,
                           data = df_panel, model = "pooling")
tanf_cases_controls <- plm(tanf_one_parent_per1000_women_1544 ~ policy_group +
                             state_unemployment_rate + minimum_wage + CPI +
                             medicaid_abortion_funding_restrictions +
                             infant_child_care_cost_burden + RPP_all_items,
                           data = df_panel, model = "pooling")
tanf_cases_twfe     <- plm(tanf_one_parent_per1000_women_1544 ~ policy_group +
                             state_unemployment_rate + minimum_wage + CPI +
                             medicaid_abortion_funding_restrictions +
                             infant_child_care_cost_burden + RPP_all_items,
                           data = df_panel, model = "within", effect = "twoways")

# SNAP
snap_ols      <- plm(snap_participation_rate ~ policy_group,
                     data = df_panel, model = "pooling")
snap_controls <- plm(snap_participation_rate ~ policy_group +
                       state_unemployment_rate + minimum_wage + CPI +
                       medicaid_abortion_funding_restrictions +
                       infant_child_care_cost_burden + RPP_all_items,
                     data = df_panel, model = "pooling")
snap_twfe     <- plm(snap_participation_rate ~ policy_group +
                       state_unemployment_rate + minimum_wage + CPI +
                       medicaid_abortion_funding_restrictions +
                       infant_child_care_cost_burden + RPP_all_items,
                     data = df_panel, model = "within", effect = "twoways")

# Abortion Rate
abort_ols      <- plm(abortions_per_1000_women_ages_1544 ~ policy_group,
                      data = df_panel, model = "pooling")
abort_controls <- plm(abortions_per_1000_women_ages_1544 ~ policy_group +
                        state_unemployment_rate + minimum_wage + CPI +
                        medicaid_abortion_funding_restrictions +
                        infant_child_care_cost_burden + RPP_all_items,
                      data = df_panel, model = "pooling")
abort_twfe     <- plm(abortions_per_1000_women_ages_1544 ~ policy_group +
                        state_unemployment_rate + minimum_wage + CPI +
                        medicaid_abortion_funding_restrictions +
                        infant_child_care_cost_burden + RPP_all_items,
                      data = df_panel, model = "within", effect = "twoways")

# Verify all models exist
exists("tanf_apps_ols")
exists("tanf_cases_ols")
exists("snap_ols")
exists("abort_ols")
```

```{r full_twfe_model}

# TANF Applications
modelsummary(list("Pooled OLS" = tanf_apps_ols,
                  "With Controls" = tanf_apps_controls,
                  "Full TWFE" = tanf_apps_twfe),
             title = "TANF Applications per 1,000 Women (15-44)",
             stars = TRUE,
             gof_map = c("nobs", "r.squared", "adj.r.squared"),
             output = "kableExtra") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = TRUE)

# TANF Caseloads
modelsummary(list("Pooled OLS" = tanf_cases_ols,
                  "With Controls" = tanf_cases_controls,
                  "Full TWFE" = tanf_cases_twfe),
             title = "TANF Caseloads per 1,000 Women (15-44)",
             stars = TRUE,
             gof_map = c("nobs", "r.squared", "adj.r.squared"),
             output = "kableExtra") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = TRUE)

# SNAP Participation
modelsummary(list("Pooled OLS" = snap_ols,
                  "With Controls" = snap_controls,
                  "Full TWFE" = snap_twfe),
             title = "SNAP Participation per 1,000 Women (15-44)",
             stars = TRUE,
             gof_map = c("nobs", "r.squared", "adj.r.squared"),
             output = "kableExtra") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = TRUE)

# Abortion Rate
modelsummary(list("Pooled OLS" = abort_ols,
                  "With Controls" = abort_controls,
                  "Full TWFE" = abort_twfe),
             title = "Abortion Rate per 1,000 Women (15-44)",
             stars = TRUE,
             gof_map = c("nobs", "r.squared", "adj.r.squared"),
             output = "kableExtra") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = TRUE)
```

### Event study plots

```{r event_study}

# Step 1: Prepare data for event study
df_event <- df_all %>%
  filter(!is.na(policy_group)) %>%
  mutate(
    treated = ifelse(policy_group == "Total Ban", 1, 0),
    event_time_binned = case_when(
      months_since_dobbs < -2 ~ -2,
      months_since_dobbs > 24 ~ 24,
      TRUE ~ as.numeric(months_since_dobbs)
    )
  )

# ============================================================
# EVENT STUDY FUNCTION
# ============================================================
run_event_study <- function(outcome_var, title, ylab, color) {
  
  formula <- as.formula(paste0(outcome_var, " ~
    i(event_time_binned, treated, ref = -1) +
    state_unemployment_rate +
    minimum_wage +
    CPI +
    medicaid_abortion_funding_restrictions +
    infant_child_care_cost_burden +
    RPP_all_items |
    State + year_month"))
  
  model <- feols(formula, data = df_event, cluster = ~State)
  
  coefs <- broom::tidy(model, conf.int = TRUE) %>%
    filter(grepl("event_time_binned::", term)) %>%
    mutate(event_time = as.numeric(gsub("event_time_binned::([-0-9]+):treated",
                                        "\\1", term)))
  
  ref_row <- data.frame(
    term = "reference", estimate = 0, std.error = 0,
    conf.low = 0, conf.high = 0, event_time = -1
  )
  
  coefs <- bind_rows(coefs, ref_row) %>% arrange(event_time)
  
  ggplot(coefs, aes(x = event_time, y = estimate)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_vline(xintercept = -0.5, linetype = "dashed", color = "red",
               linewidth = 0.8) +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
                alpha = 0.2, fill = color) +
    geom_line(color = color, linewidth = 1) +
    geom_point(color = color, size = 2.5) +
    annotate("text", x = 0, y = Inf, label = "Dobbs Decision",
             hjust = -0.1, vjust = 2, color = "red", size = 3.5) +
    scale_x_continuous(breaks = seq(-2, 24, by = 2)) +
    labs(
      title    = title,
      subtitle = "TWFE Estimates Relative to Month Before Dobbs (t = -1)",
      x        = "Months Since Dobbs (0 = June 2022)",
      y        = ylab,
      caption  = "Note: Shaded region represents 95% confidence interval. Standard errors clustered at state level."
    ) +
    theme_minimal() +
    theme(
      plot.title       = element_text(size = 13, face = "bold"),
      plot.subtitle    = element_text(size = 10, color = "gray40"),
      plot.caption     = element_text(size = 8,  color = "gray50"),
      plot.margin      = margin(10, 10, 10, 40),
      panel.grid.minor = element_blank()
    )
}

# ============================================================
# RUN ALL 4 EVENT STUDIES
# ============================================================





state_min <- as.vector(tapply(df_all$gestational_limit_weeks, df_all$State, min))
names(state_min) <- names(tapply(df_all$gestational_limit_weeks, df_all$State, min))

df_event <- df_all %>%
  filter(!is.na(policy_group)) %>%
  mutate(
    treated = if_else(as.character(policy_group) == "Total Ban", 1, 0),
    event_time_binned = as.numeric(case_when(
      months_since_dobbs < -2 ~ -2,
      months_since_dobbs > 24 ~ 24,
      TRUE ~ as.numeric(months_since_dobbs)
    ))
  )

# Verify
class(df_event$event_time_binned)
is.vector(df_event$event_time_binned)

run_event_study(
  outcome_var = "abortions_per_1000_women_ages_1544",
  title       = "Event Study: Effect of Abortion Restrictions on Abortion Rate",
  ylab        = "Estimated Effect on Abortion Rate per 1,000 Women (15-44)",
  color       = "#d73027"
)

run_event_study(
  outcome_var = "tanf._apps_per_1000",
  title       = "Event Study: Effect of Abortion Restrictions on TANF Applications",
  ylab        = "Estimated Effect on TANF Applications per 1,000 Women (15-44)",
  color       = "#fc8d59"
)

run_event_study(
  outcome_var = "tanf_one_parent_per1000_women_1544",
  title       = "Event Study: Effect of Abortion Restrictions on TANF Caseloads",
  ylab        = "Estimated Effect on TANF Caseloads per 1,000 Women (15-44)",
  color       = "#4575b4"
)

run_event_study(
  outcome_var = "snap_participation_rate",
  title       = "Event Study: Effect of Abortion Restrictions on SNAP Participation",
  ylab        = "Estimated Effect on SNAP Participation Rate",
  color       = "#1a9850"
)
```
