---
title: "Seminar preliminary tests"
author: "Julie Wilson"
format: html
editor: visual
---

# Introduction

## Research Question

Did state level abortion restrictions following Dobbs v. Jackson decision (June 2022) effect economic outcomes for women?

## Preparation

### Load Packages

```{r load_packages}


library(dplyr)
library(infer)
library(ggplot2)
library(fixest)
library(car)
library(plm)
library(readr)
install.packages("kableExtra")
install.packages("broom")
library(broom)

install.packages("corrplot")
library(corrplot)
```

# Break Down of Variables

-   dependent variables, not included together in a regression to avoid multicollinearity, each will run as a separate outcome model.

    -   Abortion_Rate - per 1000 women age 15-44, expected to decrease post-Dobbs

    -   Birth_Rate - per 1000 women age 15-44, expected to increase post-Dobbs

-   Telehealth_share

    -   share of total abortion completed via telehealth, monthly, state-level. women in ban states maintaining partial access to abortion through shield law states.

-   Medicaid_abortion_funding_restrictions

    -   1 - state restricts Medicaid funding for abortion

    -   0 - otherwise

-   Medicaid_expansion_status

    -   1 - state expanded Medicaid under ACA

    -   0 - otherwise

-   Shield_law_state

    -   1 - state enacted a shield law protecting abortion providers from out-of-state prosecution

    -   0 - otherwise

-   Shield_law_access

    -   1 - state shield law explicity protects patient access to out-of-state Telehealth abortion services

    -   0 - otherwise

-   Gestational_limit_weeks

    -   Showing point in pregnancy each state prohibits abortion (0 - total ban, 6 week, 15 week, 20 week, 24 week)

-   Single_mothers_median_monthly_income

    -   will most likely be dropped as it was used to figure stats for other variables to avoid multicollinearity.

-   Infant_childcare_cost_burden

    -   percentage of income spent on childcare as a share of single mothers median monthly income.

-   SNAP_participation_rate

    -   share of the total population enrolled in SNAP

-   TANF_caseloads_per_1000_women

    -   single parent TANF caseloads per 1000 women aged 15-44

-   State_unemployment_rate

    -   monthly state-level unemployment rate

-   Minimum_wage

    -   state level minimum wage, updated on known legislative change dates

-   CPI

    -   monthly CPI to control for inflation trends

-   RPP_all_items

    -   Regional Price Parities, controls for structural cost of living differences across states

-   Census_region

    -   categorical variable identifying each state's Census region (Northeast, South, Midwest, West) used to assign appropriate region CPI data.

### Data Import

```{r data_import}

df_prelim <- read.csv("final_data.csv")

glimpse(df_prelim)
# snap participation data did not join correctly so here I am literally spending an hour trying to merge a new data set LOL

getwd()

list.files("C:/Users/julie/Downloads/Seminar research/more r studio/Seminar research/Seminar_Preliminary")


snap_data <- read.csv("snap_participation_rate.csv")

df_prelim <- df_prelim %>%
  left_join(snap_data, by = c("State", "year_month"))


summary(df_prelim$snap_participation_rate.y)

getwd()
```

## Check for Missing Data

```{r}

sum(is.na(df_prelim))

colSums(is.na(df_prelim))

df_missing_data <- colSums(is.na(df_prelim))

which(is.na(df_prelim), arr.ind = TRUE)

df_prelim <-  df_prelim[-1651, ]

sum(is.na(df_prelim))
```

#### Missing Data Findings

Row 1651 contained missing data and was removed. leaving 1,650 complete observations. Data is now fully clean, no missing variables.

## Fixing Minimum Wage

```{r}
head(df_prelim$minimum_wage)

df_prelim$minimum_wage <- as.numeric(gsub("[^0-9.]", "", df_prelim$minimum_wage))


sum(is.na(df_prelim$minimum_wage))

which(is.na(df_prelim$minimum_wage))


sum(is.na(df_prelim$minimum_wage))
```

## Creating ban_state variable

```{r}

names(df_prelim)

df_prelim$ban_state <- ifelse(df_prelim$gestational_limit_weeks == 0, 1, 0)

table(df_prelim$ban_state)



```

# Descriptive Statistics

```{r select_data}
desc_stats <- df_analysis %>%
  filter(!is.na(policy_group)) %>%
  mutate(period = if_else(months_since_dobbs < 0,
                          "Pre-Dobbs",
                          "Post-Dobbs")) %>%
  group_by(policy_group, period) %>%
  summarize(
    mean_abortion_rate  = mean(abortions_per_1000_women_ages_1544,
                                na.rm = TRUE),
    mean_birth_rate     = mean(birth_rate_per_1000_women_1544,
                                na.rm = TRUE),
    mean_tanf           = mean(tanf_one_parent_per1000_women_1544,
                                na.rm = TRUE),
    mean_unemployment   = mean(state_unemployment_rate,
                                na.rm = TRUE),
    mean_min_wage       = mean(minimum_wage,
                                na.rm = TRUE),
    mean_medicaid       = mean(medicaid_expansion_status,
                                na.rm = TRUE),
    mean_childcare      = mean(infant_child_care_cost_burden,
                                na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  arrange(policy_group, period)

knitr::kable(desc_stats,
  col.names = c("Policy Group", "Period",
                "Abortion Rate", "Birth Rate",
                "TANF Caseloads", "Unemployment Rate",
                "Minimum Wage", "Medicaid Expansion",
                "Childcare Burden"),
  caption = "Descriptive Statistics by Policy Group Pre and Post Dobbs")
```

#### 

# T Test

```{r t_test}

t.test(df_prelim$abortions_per_1000_women_ages_1544 ~ treatment_group, data= df_pre_period)
```

This graph is showing abortion rate by policy groups.

```{r}
df_prelim <- df_prelim %>%
  mutate(policy_group = case_when(
    shield_law_state == 1 ~ "Shield Law (Protective)",
    gestational_limit_weeks <= 12 ~ "Restrictive",
    is.na(gestational_limit_weeks) ~ "Restrictive",
    TRUE ~ "Middle Ground"
    ),
    shield_group = if_else(shield_law_state == 1, 
                           "Shield Law State", "No Shield Law")
  )

df_prelim %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(mean_abortion = mean(abortions_per_1000_women_ages_1544, na.rm = TRUE), groups = "drop") %>%
  ggplot(aes(x = months_since_dobbs, y = mean_abortion, 
             color = policy_group)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "pink") +
  annotate("text", x = 0.5, y = Inf, label = "Dobbs Decision", 
           hjust = 0, vjust = 2, color = "pink", size = 3.5) +
  labs(title = "Abortion Rate by State Policy Group",
       x = "Months Since Dobbs (0 = June 2022)",
       y = "Abortions per 1,000 Women (15-44)",
       color = "Policy Group") +
  theme_minimal()
```

## Testing for Multicollinearity

```{r}

#  removing gestational limit weeks due to multicollinearity.

model_vif <- lm(abortions_per_1000_women_ages_1544 ~ ban_state + 
                shield_law_state + shield_law_access +
                medicaid_expansion_status + 
                medicaid_abortion_funding_restrictions +
                minimum_wage + CPI + RPP_all_items + 
                state_unemployment_rate,
                data = df_prelim)


```

#### Multicollinearity (variance inflation factor) findings

-   Ban_state has a VIF of 7.12

-   gestational_limit_weeks has a VIF of 8.09

I am going to remove the gestational limit variable from the regression model and keep ban state to keep VIF clean.

-   Now there is no multicollinearity among predictor variables after removing gestational limit weeks.

## Parallel Trends Credibility Check

-   Fundamental identification assumption for Difference-in-Differences (DiD) model, which posits that in the absence of the *Dobbs* decision, treatment and control states would have followed similar economic trajectories.

```{r}
# ============================================================
# DATA CLEANING AND PREP
# ============================================================

# Fix character variables
df_prelim$single_mothers_median_monthly_income <- as.numeric(gsub(
  "[$,]", "", df_prelim$single_mothers_median_monthly_income))
df_prelim$infant_child_care_cost_burden <- as.numeric(gsub(
  "%", "", df_prelim$infant_child_care_cost_burden))

# Fix SNAP join - drop the old snap column but keep everything else
df_prelim <- df_prelim %>%
  select(-snap_participation_rate.y) %>%
  left_join(snap_data, by = c("State", "year_month"))

# Drop rows with missing State or months_since_dobbs
df_prelim <- df_prelim %>%
  filter(!is.na(State),
         !is.na(months_since_dobbs))

# Confirm range and structure look correct
range(df_prelim$months_since_dobbs)
nrow(df_prelim)
names(df_prelim)
head(df_prelim)

head(df_prelim$months_since_dobbs, 20)

df_prelim <- df_prelim %>%
  filter(!is.na(State))

nrow(df_prelim)

range(df_prelim$months_since_dobbs)

df_prelim %>%
  filter(is.na(months_since_dobbs)) %>%
  head()

df_prelim <- df_prelim %>%
  filter(!is.na(months_since_dobbs))

nrow(df_prelim)
range(df_prelim$months_since_dobbs)
```

```{r}
# ============================================================
# POLICY GROUP CLASSIFICATION BASED ON POST-DOBBS BEHAVIOR
# ============================================================

# Step 1: Identify states that enacted restrictions post-Dobbs
# These are states whose minimum gestational limit is 0 or 6
# meaning they enacted a ban or severe restriction after Dobbs

restrictive_states <- c(
  # Total ban states
  "AL", "AR", "ID", "IN", "KY", "LA", "MO", 
  "MS", "ND", "OK", "SD", "TN", "TX", "WV", "WY",
  # 6 week states
  "FL", "GA", "IA", "SC"
)

non_restrictive_states <- c(
  "AK", "CA", "CO", "CT", "DE", "HI", "IL", 
  "KS", "ME", "MD", "MA", "MI", "MN", "MT", 
  "NV", "NH", "NJ", "NM", "NY", "OR", "PA", 
  "RI", "VT", "VA", "WA"
)

# Step 2: Classify states
df_prelim <- df_prelim %>%
  mutate(policy_group = case_when(
    State %in% restrictive_states     ~ "Restrictive",
    State %in% non_restrictive_states ~ "Non-Restrictive",
    TRUE                              ~ NA_character_
  ))

# Step 3: Drop middle ground states
df_analysis <- df_prelim %>%
  filter(!is.na(policy_group))

# Step 4: Confirm classification
df_analysis %>%
  distinct(State, policy_group) %>%
  arrange(policy_group, State) %>%
  as.data.frame() %>%
  print()

# Step 5: Check state counts
df_analysis %>%
  distinct(State, policy_group) %>%
  count(policy_group)

# Step 6: Check observation count
nrow(df_analysis)
table(df_analysis$policy_group)
```

```{r}
# ============================================================
# PARALLEL TRENDS CHECK - UPDATED CLASSIFICATION
# ============================================================

df_parallel <- df_analysis %>%
  filter(!is.na(policy_group),
         months_since_dobbs >= -3,
         months_since_dobbs <= 0) %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(
    mean_tanf = mean(tanf_one_parent_per1000_women_1544, na.rm = TRUE),
    se_tanf   = sd(tanf_one_parent_per1000_women_1544, na.rm = TRUE) /
                sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(
    ci_low  = mean_tanf - 1.96 * se_tanf,
    ci_high = mean_tanf + 1.96 * se_tanf
  )

ggplot(df_parallel,
       aes(x = months_since_dobbs, y = mean_tanf,
           color = policy_group, fill = policy_group)) +
  geom_ribbon(aes(ymin = ci_low, ymax = ci_high),
              alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0,
             linetype = "dashed",
             color = "red",
             linewidth = 0.8) +
  annotate("text",
           x = -0.1, y = Inf,
           label = "Dobbs Decision",
           hjust = 1, vjust = 2,
           color = "red",
           size = 3.5) +
  scale_color_manual(values = c(
    "Restrictive"     = "#d73027",
    "Non-Restrictive" = "#1a9850"
  )) +
  scale_fill_manual(values = c(
    "Restrictive"     = "#d73027",
    "Non-Restrictive" = "#1a9850"
  )) +
  scale_x_continuous(breaks = c(-3, -2, -1, 0)) +
  labs(
    title    = "Parallel Trends Check - Pre-Dobbs TANF Caseloads",
    x        = "Months Since Dobbs (0 = June 2022)",
    y        = "TANF Cases per 1,000 Women (15-44)",
    color    = "Policy Group",
    fill     = "Policy Group"
  ) +
  theme_minimal() +
  theme(plot.margin = margin(10, 10, 10, 40))
```

```{r}
df_analysis %>%
  filter(!is.na(policy_group),
         months_since_dobbs >= -3,
         months_since_dobbs <= 0) %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(mean_tanf = mean(tanf_one_parent_per1000_women_1544,
                              na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = months_since_dobbs, y = mean_tanf,
             color = policy_group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = -0.1, y = Inf, label = "Dobbs Decision",
           hjust = 1, vjust = 2, color = "red", size = 3.5) +
  scale_color_manual(values = c(
    "Restrictive"     = "#d73027",
    "Non-Restrictive" = "#1a9850"
  )) +
  scale_x_continuous(breaks = c(-3, -2, -1, 0)) +
  labs(title = "Parallel Trends Check - Pre-Dobbs TANF Caseloads",
       x = "Months Since Dobbs (0 = June 2022)",
       y = "TANF Cases per 1,000 Women (15-44)",
       color = "Policy Group") +
  theme_minimal() +
  theme(plot.margin = margin(10, 10, 10, 40))


```

```{r}
# Fix character variables
df_prelim$single_mothers_median_monthly_income <- as.numeric(gsub(
  "[$,]", "", df_prelim$single_mothers_median_monthly_income))
df_prelim$infant_child_care_cost_burden <- as.numeric(gsub(
  "%", "", df_prelim$infant_child_care_cost_burden))

# Fix SNAP join
df_prelim <- df_prelim %>%
  select(-snap_participation_rate) %>%
  left_join(snap_data, by = c("State", "year_month"))

# Identify restrictive states by abortion rate dropping to 0
restrictive_states <- df_prelim %>%
  filter(abortions_per_1000_women_ages_1544 == 0) %>%
  distinct(State) %>%
  pull(State)

# Recreate policy group
df_prelim <- df_prelim %>%
  mutate(policy_group = case_when(
    State %in% restrictive_states ~ "Restrictive",
    TRUE                          ~ "Non-Restrictive"
  ))

# Confirm range is correct
range(df_prelim$months_since_dobbs)
head(df_prelim)
names(df_prelim)
nrow(df_prelim)
head(df_prelim$months_since_dobbs, 20)
df_prelim <- df_prelim %>%
  filter(!is.na(State))
nrow(df_prelim)
range(df_prelim$months_since_dobbs)
df_prelim %>%
  filter(is.na(months_since_dobbs)) %>%
  head()
df_prelim <- df_prelim %>%
  filter(!is.na(months_since_dobbs))
nrow(df_prelim)
range(df_prelim$months_since_dobbs)

```

### TANF caseload by policy group, pre- and post- Dobbs.

```{r}

getwd()

df_prelim %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(mean_tanf = mean(tanf_one_parent_per1000_women_1544,
                              na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = months_since_dobbs, y = mean_tanf,
             color = policy_group)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = 0.5, y = Inf, label = "Dobbs Decision",
           hjust = 0, vjust = 2, color = "red", size = 3.5) +
  labs(title = "TANF Caseloads by State Policy Group - Pre and Post Dobbs",
       x = "Months Since Dobbs (0 = June 2022)",
       y = "TANF Cases per 1,000 Women (15-44)",
       color = "Policy Group") +
  theme_minimal()

```

#### Parallel Trends Credibility Check results:

-   Blue (shield law states)

-   Green (restrictive)

-   Red (middle ground)

-   

### SNAP participation by policy group, pre- and post- Dobbs.

```{r}
df_prelim %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(mean_snap = mean(snap_participation_rate,
                              na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = months_since_dobbs, y = mean_snap,
             color = policy_group)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = 0.5, y = Inf, label = "Dobbs Decision",
           hjust = 0, vjust = 2, color = "red", size = 3.5) +
  labs(title = "SNAP Participation by State Policy Group - Pre and Post Dobbs",
       x = "Months Since Dobbs (0 = June 2022)",
       y = "SNAP Participation Rate",
       color = "Policy Group") +
  theme_minimal()
```

#### Parallel Trends Credibility Check results:

```{r}
# ============================================================
# POLICY GROUP CLASSIFICATION
# ============================================================

# Step 1: Classify states by gestational limit
# Restrictive = total ban (0) or ≤ 6 weeks
# Non-Restrictive = 20+ weeks or no stated limit
# Middle ground states are excluded from analysis sample

df_prelim <- df_prelim %>%
  mutate(policy_group = case_when(
    gestational_limit_weeks == 0  ~ "Restrictive",
    gestational_limit_weeks <= 6  ~ "Restrictive",
    gestational_limit_weeks >= 20 ~ "Non-Restrictive",
    TRUE                          ~ NA_character_  # excludes middle ground
  ))

# Step 2: Drop middle ground states from analysis sample
df_analysis <- df_prelim %>%
  filter(!is.na(policy_group))

# Step 3: Confirm classification looks correct
df_analysis %>%
  distinct(State, policy_group, gestational_limit_weeks) %>%
  arrange(policy_group, gestational_limit_weeks) %>%
  as.data.frame() %>%
  print()

# Step 4: Check how many states fall in each group
df_analysis %>%
  distinct(State, policy_group) %>%
  count(policy_group)

# Step 5: Check observation count after dropping middle ground
nrow(df_analysis)

# ============================================================
# RESIDUALIZE SNAP AND TANF FOR CPI AND RPP
# ============================================================

df_analysis <- df_analysis %>%
  mutate(
    tanf_resid = residuals(lm(tanf_one_parent_per1000_women_1544 ~ CPI + RPP_all_items,
                               data = .)),
    snap_resid = residuals(lm(snap_participation_rate ~ CPI + RPP_all_items,
                               data = .))
  )

# ============================================================
# CONFIRM FINAL DATASET LOOKS CORRECT
# ============================================================

range(df_analysis$months_since_dobbs)
nrow(df_analysis)
names(df_analysis)
head(df_analysis)

# Confirm no missing months_since_dobbs
df_analysis %>%
  filter(is.na(months_since_dobbs)) %>%
  nrow()

# Final clean - drop any remaining NA rows in key columns
df_analysis <- df_analysis %>%
  filter(!is.na(State),
         !is.na(months_since_dobbs),
         !is.na(policy_group))

# Final confirmation
nrow(df_analysis)
range(df_analysis$months_since_dobbs)
table(df_analysis$policy_group)

# Check what is causing the TX duplicate
df_analysis %>%
  filter(State == "TX") %>%
  distinct(State, gestational_limit_weeks, year_month) %>%
  arrange(year_month) %>%
  head(20)
```

```{r}
# ============================================================
# TANF CASELOADS GRAPH - PRE AND POST DOBBS (CPI ONLY)
# ============================================================

df_analysis %>%
  filter(!is.na(policy_group)) %>%
  mutate(tanf_adjusted = tanf_one_parent_per1000_women_1544 / CPI * 100) %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(mean_tanf = mean(tanf_adjusted, na.rm = TRUE), .groups = "drop") %>%
  group_by(policy_group) %>%
  mutate(baseline = mean(mean_tanf[months_since_dobbs < 0], na.rm = TRUE),
         mean_tanf_indexed = mean_tanf - baseline) %>%
  ggplot(aes(x = months_since_dobbs, y = mean_tanf_indexed,
             color = policy_group)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = 0.5, y = Inf, label = "Dobbs Decision",
           hjust = 0, vjust = 2, color = "red", size = 3.5) +
  scale_color_manual(values = c(
    "Restrictive"     = "#d73027",
    "Non-Restrictive" = "#1a9850"
  )) +
  labs(title = "TANF Caseloads by State Policy Group - Pre and Post Dobbs",
       x = "Months Since Dobbs (0 = June 2022)",
       y = "Change in TANF Caseloads from Pre-Dobbs Baseline (CPI Adjusted)",
       color = "Policy Group") +
  theme_minimal() +
  theme(plot.margin = margin(10, 10, 10, 40))
```

```{r}
# ============================================================
# SNAP PARTICIPATION GRAPH - PRE AND POST DOBBS (CPI ONLY)
# ============================================================

df_analysis %>%
  filter(!is.na(policy_group)) %>%
  mutate(snap_adjusted = snap_participation_rate / CPI * 100) %>%
  group_by(months_since_dobbs, policy_group) %>%
  summarize(mean_snap = mean(snap_adjusted, na.rm = TRUE), .groups = "drop") %>%
  group_by(policy_group) %>%
  mutate(baseline = mean(mean_snap[months_since_dobbs < 0], na.rm = TRUE),
         mean_snap_indexed = mean_snap - baseline) %>%
  ggplot(aes(x = months_since_dobbs, y = mean_snap_indexed,
             color = policy_group)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = 0.5, y = Inf, label = "Dobbs Decision",
           hjust = 0, vjust = 2, color = "red", size = 3.5) +
  scale_color_manual(values = c(
    "Restrictive"     = "#d73027",
    "Non-Restrictive" = "#1a9850"
  )) +
  labs(title = "SNAP Participation by State Policy Group - Pre and Post Dobbs",
       x = "Months Since Dobbs (0 = June 2022)",
       y = "Change in SNAP Participation from Pre-Dobbs Baseline (CPI Adjusted)",
       color = "Policy Group") +
  theme_minimal() +
  theme(plot.margin = margin(10, 10, 10, 40))
```

## Hausman Specification Test

##### Fixed vs Random effects model specification

estimated to determine the appropriate panel estimator for the data

In previous abortion research, a result of p\<0.05 typically indicates that state-specific effects are correlated with explanatory variables, meaning the Fixed Effects estimator is preferred to avoid bias

### Random Effects

```{r}
# Running both fixed effects and random effects to double check that Fixed Effects is what I want 

#starting wit the random effects: 
df_panel <- pdata.frame(df_prelim, index = c("State", "year_month"))

class(df_panel)

df_panel$infant_child_care_cost_burden <- as.numeric(gsub("%", "", 
                                          df_panel$infant_child_care_cost_burden))

re_model <- plm(abortions_per_1000_women_ages_1544 ~ 
                  gestational_limit_weeks +
                  shield_law_state +
                  state_unemployment_rate +
                  minimum_wage +
                  CPI +
                  RPP_all_items +
                  infant_child_care_cost_burden,
                data = df_panel,
                model = "random")
summary(re_model)
```

#### Random Effects Results:

-   gestational_limit_weeks - highly significant (p\<0.001)

    -   more restrictive limits strongly predict lower abortion rates (obviously)

-   CPI - highly significant (P\<0.001)

-   infant_child_care_cost_burden - significant (p\<0.01)

    -   has negative coefficient, meaning higher childcare costs are associated with lower abortion rates (uhhh weird??)

-   state_unemployment_rate - significant (p\<0.05)

-   minimum_wage - not significant (p=0.28)

-   RPP_all_items - not significant (p=0.47)

-   shield_law_state - borderline (p=0.056)

### Fixed Effects

```{r}
fe_model <- plm(abortions_per_1000_women_ages_1544 ~ 
                  gestational_limit_weeks +
                  shield_law_state +
                  state_unemployment_rate +
                  minimum_wage +
                  CPI +
                  RPP_all_items +
                  infant_child_care_cost_burden,
                data = df_panel,
                model = "within")

exists("fe_model")


summary(fe_model)
```

#### Fixed Effects Results:

-   RPP_all_items - not significant in RE, now significant (p\<0.05)

    -   once I control for state specific characteristics, regional price parities do matter

-   Shield_law_state - dropped from borderline to not significant (p=0.12)

    -   suggesting shield law status is partly capturing state level characteristics rather than a pure policy effect

-   Minimum_wage - remains not significant in both models

-   Variables still consistent

    -   gestational_limit_weeks

    -   CPI

    -   infant_child_care_cost_burden

    -   state_unemployment_rate

## Hausman Test

```{r}
phtest(fe_model, re_model)
```

## Strict Exogeneity Test

This helps confirm that explanatory variables are not correlated with the model's error term, ensuring estimated coefficients are unbiased

```{r}
#creating lead variables for key policy regressors 

df_panel <- df_panel %>%
  group_by(State) %>%
  arrange(year_month) %>%
  mutate(
    lead_gestational_limit = dplyr::lead(gestational_limit_weeks, 1),
    lead_shield_law = dplyr::lead(shield_law_state, 1),
    lead_shield_access = dplyr::lead(shield_law_access, 1),
    lead_medicaid_restrictions = dplyr::lead(medicaid_abortion_funding_restrictions, 1)
  ) %>%
  ungroup()

# checking it worked 

df_panel %>%
  select(State, year_month, gestational_limit_weeks, lead_gestational_limit) %>%
  head(10)


df_panel %>%
  select(State, year_month, gestational_limit_weeks, lead_gestational_limit) %>%
  filter(State == "AK") %>%
  head(10)

# alaska gestational limit remained the same all the way through the study period with a limit of 24 weeks. Now lets compare it to a state like Texas that did change throughout the study period 

df_panel %>%
  select(State, year_month, gestational_limit_weeks, lead_gestational_limit) %>%
  filter(State == "TX") %>%
  head(10)

# texas shows a gestational limit of 6 weeks before dobbs that drops to 0 after the trigger ban, the lead column correctly previews that drop one month early. this confirms the lead variable is working as intended for the exogeneity test. 
```

### Strict exogeneity test - Primary outcome: Abortion Rate with policy variables

```{r}

df_panel$single_mothers_median_monthly_income <-
readr::parse_number(as.character(df_panel$single_mothers_median_monthly_income))


df_panel$single_mothers_median_monthly_income <-
readr::parse_number(as.character(df_panel$single_mothers_median_monthly_income))

class(df_panel$single_mothers_median_monthly_income)

head(df_panel$single_mothers_median_monthly_income)

df_panel <- pdata.frame(df_panel, index = c("State", "year_month"))

names(df_panel)

df_panel <- pdata.frame(df_panel, index = c("State", "year_month"))

exo_abortion <- plm(
abortions_per_1000_women_ages_1544 ~
gestational_limit_weeks +
shield_law_state +
shield_law_access +
medicaid_abortion_funding_restrictions +
lead_gestational_limit +
lead_shield_law +
lead_shield_access +
lead_medicaid_restrictions,
data = df_panel,
model = "within"
)

summary(exo_abortion)


```

#### results

-   Both significant - future shield law changes do not predict current abortion rates. Policy variables are strictly exogenous and the estimates are reliable

    -   lead_shield_law (p=0.29)

    -   lead_shield_access (p=0.36)

## Strict Exogeneity Test - Abortion Rate with Economic Variables

Checking whether the policy and economic variables in in the model are truly independent of future outcomes, or whether abortion rates were already responding to anticipated policy changes before they officially took effect.

```{r}
exo_abortion2 <- plm(abortions_per_1000_women_ages_1544 ~
                      gestational_limit_weeks +
                      infant_child_care_cost_burden +
                      single_mothers_median_monthly_income +
                      snap_participation_rate +
                      tanf_one_parent_per1000_women_1544 +
                      lead_gestational_limit +
                      lead_shield_law +
                      lead_shield_access +
                      lead_medicaid_restrictions,
                    data = df_panel,
                    model = "within")

summary(exo_abortion2)

```

#### results

-   flagged for potential anticipation effects - women and providers may have been responding to expected policy changes before they took effect. A known limitation in abortion access research (noted as a caveat)

    -   gestational_limit_weeks

    -   shield_law_state

```{r}
# ============================================================
# TWFE EVENT STUDY - TANF CASELOADS
# ============================================================



# Step 1: Prepare data
df_event <- df_analysis %>%
  filter(!is.na(policy_group)) %>%
  mutate(
    treated = if_else(policy_group == "Restrictive", 1, 0),
    event_time_binned = case_when(
      months_since_dobbs < -2 ~ -2,
      months_since_dobbs > 24 ~ 24,
      TRUE ~ as.numeric(months_since_dobbs)
    )
  )

# Step 2: Estimate TWFE event study model
es_tanf <- feols(
  tanf_one_parent_per1000_women_1544 ~
    i(event_time_binned, treated, ref = -1) +
    state_unemployment_rate +
    minimum_wage +
    CPI +
    medicaid_expansion_status |
    State + year_month,
  data = df_event,
  cluster = ~State
)

summary(es_tanf)

# Step 3: Extract coefficients
es_coefs_tanf <- broom::tidy(es_tanf, conf.int = TRUE) %>%
  filter(grepl("event_time_binned::", term)) %>%
  mutate(
    event_time = as.numeric(
      gsub("event_time_binned::([-0-9]+):treated", "\\1", term)
    )
  )

# Step 4: Add reference period manually at t = -1
ref_row <- data.frame(
  term       = "reference",
  estimate   = 0,
  std.error  = 0,
  conf.low   = 0,
  conf.high  = 0,
  event_time = -1
)

es_coefs_tanf <- bind_rows(es_coefs_tanf, ref_row) %>%
  arrange(event_time)

# Step 5: Plot
ggplot(es_coefs_tanf,
       aes(x = event_time, y = estimate)) +
  geom_hline(yintercept = 0,
             linetype = "dashed",
             color = "gray50") +
  geom_vline(xintercept = -0.5,
             linetype = "dashed",
             color = "red",
             linewidth = 0.8) +
  geom_ribbon(aes(ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.2,
              fill = "#d73027") +
  geom_line(color = "#d73027",
            linewidth = 1) +
  geom_point(color = "#d73027",
             size = 2.5) +
  annotate("text",
           x = 0, y = Inf,
           label = "Dobbs Decision",
           hjust = -0.1, vjust = 2,
           color = "red",
           size = 3.5) +
  scale_x_continuous(breaks = seq(-2, 24, by = 2)) +
  labs(
    title    = "Event Study: Effect of Abortion Restrictions on TANF Caseloads",
    subtitle = "TWFE Estimates Relative to Month Before Dobbs (t = -1)",
    x        = "Months Since Dobbs (0 = June 2022)",
    y        = "Estimated Effect on TANF Cases per 1,000 Women (15-44)",
    caption  = "Note: Shaded region represents 95% confidence interval. Standard errors clustered at state level."
  ) +
  theme_minimal() +
  theme(
    plot.title       = element_text(size = 13, face = "bold"),
    plot.subtitle    = element_text(size = 10, color = "gray40"),
    plot.caption     = element_text(size = 8,  color = "gray50"),
    plot.margin      = margin(10, 10, 10, 40),
    panel.grid.minor = element_blank()
  )
```

```{r}
# ============================================================
# TWFE EVENT STUDY - ABORTION RATE
# ============================================================



# Step 1: Prepare data for event study
df_event <- df_analysis %>%
  filter(!is.na(policy_group)) %>%
  mutate(
    treated = if_else(policy_group == "Restrictive", 1, 0),
    # Bin endpoints to avoid sparse estimates at edges
    event_time_binned = case_when(
      months_since_dobbs < -2 ~ -2,
      months_since_dobbs > 24 ~ 24,
      TRUE ~ as.numeric(months_since_dobbs)
    )
  )

# Step 2: Estimate TWFE event study model
# ref = -1 drops the period immediately before Dobbs as baseline
es_abortion <- feols(
  abortions_per_1000_women_ages_1544 ~
    i(event_time_binned, treated, ref = -1) +
    state_unemployment_rate +
    minimum_wage +
    CPI +
    medicaid_expansion_status |
    State + year_month,
  data = df_event,
  cluster = ~State
)

summary(es_abortion)

# Step 3: Extract coefficients
es_coefs_abortion <- broom::tidy(es_abortion, conf.int = TRUE) %>%
  filter(grepl("event_time_binned::", term)) %>%
  mutate(
    event_time = as.numeric(gsub("event_time_binned::([-0-9]+):treated", 
                                  "\\1", term))
  )

# Step 4: Add reference period manually (coefficient = 0 at t = -1)
ref_row <- data.frame(
  term      = "reference",
  estimate  = 0,
  std.error = 0,
  conf.low  = 0,
  conf.high = 0,
  event_time = -1
)

es_coefs_abortion <- bind_rows(es_coefs_abortion, ref_row) %>%
  arrange(event_time)

# Step 5: Plot
ggplot(es_coefs_abortion, 
       aes(x = event_time, y = estimate)) +
  geom_hline(yintercept = 0, 
             linetype = "dashed", 
             color = "gray50") +
  geom_vline(xintercept = -0.5, 
             linetype = "dashed", 
             color = "red",
             linewidth = 0.8) +
  geom_ribbon(aes(ymin = conf.low, 
                  ymax = conf.high),
              alpha = 0.2,
              fill = "#d73027") +
  geom_line(color = "#d73027", 
            linewidth = 1) +
  geom_point(color = "#d73027", 
             size = 2.5) +
  annotate("text", 
           x = 0, y = Inf, 
           label = "Dobbs Decision",
           hjust = -0.1, vjust = 2, 
           color = "red", 
           size = 3.5) +
  scale_x_continuous(breaks = seq(-2, 24, by = 2)) +
  labs(
    title    = "Event Study: Effect of Abortion Restrictions on Abortion Rate",
    subtitle = "TWFE Estimates Relative to Month Before Dobbs (t = -1)",
    x        = "Months Since Dobbs (0 = June 2022)",
    y        = "Estimated Effect on Abortion Rate per 1,000 Women (15-44)",
    caption  = "Note: Shaded region represents 95% confidence interval. Standard errors clustered at state level."
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(size = 10, color = "gray40"),
    plot.caption  = element_text(size = 8,  color = "gray50"),
    plot.margin   = margin(10, 10, 10, 40),
    panel.grid.minor = element_blank()
  )
```

```{r}
# ============================================================
# BASELINE OLS REGRESSION TABLE
# ============================================================

library(modelsummary)

# Model 1 - Simple OLS no controls
model_ols_simple_tanf <- feols(
  tanf_one_parent_per1000_women_1544~
    treated,
  data = df_event
)

model_ols_simple_abortion <- feols(
  abortions_per_1000_women_ages_1544 ~
    treated,
  data = df_event
)

# Model 2 - OLS with economic controls
model_ols_controls_tanf <- feols(
  tanf_one_parent_per1000_women_1544 ~
    treated +
    state_unemployment_rate +
    minimum_wage +
    CPI,
  data = df_event
)

model_ols_controls_abortion <- feols(
  abortions_per_1000_women_ages_1544 ~
    treated +
    state_unemployment_rate +
    minimum_wage +
    CPI,
  data = df_event
)

# Model 3 - TWFE with fixed effects and controls
model_twfe_tanf <- feols(
  tanf_one_parent_per1000_women_1544 ~
    treated +
    state_unemployment_rate +
    minimum_wage +
    CPI |
    State + year_month,
  data = df_event,
  cluster = ~State
)

model_twfe_abortion <- feols(
  abortions_per_1000_women_ages_1544 ~
    treated +
    state_unemployment_rate +
    minimum_wage +
    CPI |
    State + year_month,
  data = df_event,
  cluster = ~State
)

# ============================================================
# COMBINED TABLE - TANF OUTCOME
# ============================================================

modelsummary(
  list(
    "Pooled OLS"     = model_ols_simple_tanf,
    "OLS + Controls" = model_ols_controls_tanf,
    "TWFE"           = model_twfe_tanf
  ),
  stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
  coef_map = c(
    "treated"                 = "Restrictive State (Treated)",
    "state_unemployment_rate" = "Unemployment Rate",
    "minimum_wage"            = "Minimum Wage",
    "CPI"                     = "CPI"
  ),
  gof_map = list(
    list(raw = "nobs",      clean = "Observations", fmt = 0),
    list(raw = "adj.r.squared", clean = "Adj. R²",           fmt = 3)
  ),
  title  = "Regression Results: Effect of Abortion Restrictions on TANF Caseloads",
  notes  = "Columns 1-2 are pooled OLS. Column 3 uses TWFE with state and time fixed effects. Standard errors clustered at state level for TWFE. * p<0.1, ** p<0.05, *** p<0.01",
  output = "gt"
)


```

```{r}
# Pooled OLS - no controls
model_ols_simple_abortion <- feols(
  abortions_per_1000_women_ages_1544 ~
    treated,
  data = df_event
)

# OLS with controls - keep medicaid restriction
model_ols_controls_abortion <- feols(
  abortions_per_1000_women_ages_1544 ~
    treated +
    state_unemployment_rate +
    minimum_wage +
    CPI +
    medicaid_abortion_funding_restrictions,
  data = df_event
)

# TWFE - drop medicaid restriction and treated since fixed effects absorb them
model_twfe_abortion <- feols(
  abortions_per_1000_women_ages_1544 ~
    state_unemployment_rate +
    minimum_wage +
    CPI |
    State + year_month,
  data = df_event,
  cluster = ~State
)

# Confirm no collinearity warnings
summary(model_twfe_abortion)

modelsummary(
  list(
    "Pooled OLS"     = model_ols_simple_abortion,
    "OLS + Controls" = model_ols_controls_abortion,
    "TWFE"           = model_twfe_abortion
  ),
  stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
  coef_map = c(
    "treated"                               = "Restrictive State (Treated)",
    "state_unemployment_rate"               = "Unemployment Rate",
    "minimum_wage"                          = "Minimum Wage",
    "CPI"                                   = "CPI",
    "medicaid_abortion_funding_restrictions" = "Medicaid Abortion Funding Restrictions"
  ),
  gof_map = list(
    list(raw = "nobs",          clean = "Observations", fmt = 0),
    list(raw = "adj.r.squared", clean = "Adj. R²",      fmt = 3)
  ),
  title  = "Regression Results: Effect of Abortion Restrictions on Abortion Rate",
  notes  = "Columns 1-2 are pooled OLS. Column 3 uses TWFE with state and time fixed effects. Standard errors clustered at state level for TWFE. * p<0.1, ** p<0.05, *** p<0.01",
  output = "gt"
)
```

```{r}
modelsummary(
  list(
    "Pooled OLS"     = model_ols_simple_abortion,
    "OLS + Controls" = model_ols_controls_abortion,
    "TWFE"           = model_twfe_abortion
  ),
  stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
  coef_map = c(
    "treated"                              = "Restrictive State (Treated)",
    "state_unemployment_rate"              = "Unemployment Rate",
    "minimum_wage"                         = "Minimum Wage",
    "CPI"                                  = "CPI",
    "medicaid_abortion_funding_restrictions" = "Medicaid Abortion Funding Restrictions",
    "telehealth_share"                     = "Telehealth Share"
  ),
  gof_map = list(
    list(raw = "nobs",          clean = "Observations", fmt = 0),
    list(raw = "adj.r.squared", clean = "Adj. R²",      fmt = 3)
  ),
  title  = "Regression Results: Effect of Abortion Restrictions on Abortion Rate",
  notes  = "Columns 1-2 are pooled OLS. Column 3 uses TWFE with state and time fixed effects. Standard errors clustered at state level for TWFE. * p<0.1, ** p<0.05, *** p<0.01",
  output = "gt"
)

# Check what coefficients are in the updated models
names(coef(model_ols_controls_abortion))
names(coef(model_twfe_abortion))
```

AI Disclosure: AI was used to help build codes throughout the entire assignment.

Also, this is a mess... I know.... itll be better when I start from scratch and redo my policy groups and models.
